### 2022-03-04 paper/localpsf.tex
add back some details to RBF section

%The ellipsoid $E_x$ is not immediately available. 
To determine $E_x$, we must compute the scalar $\spatialvol(x)$, the length-$\gdim$ vector $\spatialmean(x)$, and the $\gdim \times \gdim$ matrix $\spatialcov(x)$.
%(recall $\gdim$ is the spatial dimension of $\Omega$).
The direct approach to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$. But this direct approach is computationally infeasible because it requires one operator action of $\Aop$ per point $x$ that we wish to determine $E_x$ at, and we will need to determine $E_x$ at a large number of points. Fortunately, it is possible to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$, \emph{for all points $x\in\Omega$ simultaneously}, by applying $\Aop^T$ to one constant function, $\gdim$ linear functions, and $\gdim(\gdim+1)/2$ quadratic functions. We present this efficient method of computing $\spatialvol$, $\spatialmean$, and $\spatialcov$ in Theorem \ref{thm:vol_mean_cov} and summarize the method for computing these moments in Algorithm \ref{alg:varhpi_mean_cov}.

\begin{theorem}[Impulse response moments]
	\label{thm:vol_mean_cov}
%	Let $\spatialvol:\Omega \rightarrow \mathbb{R}$, $\spatialmean:\Omega \rightarrow \mathbb{R}^\gdim$, and $\spatialcov:\Omega \rightarrow \mathbb{R}^{\gdim \times \gdim}$, be the following functions:
%	\begin{equation*}
%	\spatialvol(x) := \int_{\Omega} \impulseresponse_x(y) dy, \qquad
%	\spatialmean(x) := \mathbb{E}(\widehat{\impulseresponse}_x), \qquad
%	\spatialcov(x) := \Var(\widehat{\impulseresponse}_x),
%	\end{equation*}
%	where $\widehat{\impulseresponse}_x := \impulseresponse_x / \spatialvol(x)$ denotes the normalized version of $\impulseresponse_x$, $\mathbb{E}(\widehat{\impulseresponse}_x)$ denotes the mean of $\widehat{\impulseresponse}_x$, and $\Var(\widehat{\impulseresponse}_x)$ denotes the variance of $\widehat{\impulseresponse}_x$.	
	Let  $x^i$ denote the $i^\text{th}$ component of $x$, i.e., $x = \left(x^1, x^2, \dots, x^\gdim\right)$. Let $C$, $\{L^i\}_{i=1}^\gdim$, and ${\{Q^{ij}\}_{i=1}^\gdim}_{j=1}^\gdim$ be the following constant, linear, and quadratic functions:
	\begin{equation*}
	C(x) := 1, \qquad
	L^i(x) := x^i, \qquad
	Q^{ij}(x) = x^i x^j.
	\end{equation*}
	Let $f/g$ denote the pointwise division of functions, $\left(f/g\right)(x) = f(x)/g(x)$, and $f\cdot g$ denote the pointwise multiplication of functions, $(f\cdot g)(x) = f(x)g(x)$. The following results hold:
	\begin{subequations}
		\label{eq:vol_mean_var}
		\begin{align}
		\spatialvol =& \left(\Aop^T C\right)^* \\
		\spatialmean^i =& \left(\Aop^T L^i\right)^* / \spatialvol \\
		\spatialcov^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j
		\end{align}
	\end{subequations}
	for $i=1,\dots, \gdim$, $j=1,\dots,\gdim$.
\end{theorem}

The proof of Theorem \ref{thm:vol_mean_cov} is shown in Appendix \ref{app:proofs}. The proof is straightforward, and follows directly from the definitions.

@@

%In Section \ref{sec:intro} we noted that computing $\Aker\left(y,x\right)$ is computationally costly. 
%This section describes how we compute approximate kernel entries
%\begin{equation*}
%\widetilde{\Phi}(y,x) \approx \Phi(y,x)
%\end{equation*}
%for arbitrary pairs of points $(y,x) \in \Omega \times \Omega$. 
Given $(y,x)\in \Omega \times \Omega$, let
\begin{equation}
\label{eq:fxyxp}
f_i := \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right), \qquad i=1,\dots,k,
\end{equation}
%for arbitrary pairs of points $(y,x) \subset \Omega \times \Omega$.
% while Section \ref{sec:approximate_kernel_entries} we showed how we can cheaply evaluate approximate kernel entries at arbitrary pairs of points $(y,x) \in \Omega \times \Omega$.
where $\{x_i\}_{i=1}^k$ are the $k$ nearest sample points to $x$, excluding sample points for which $\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right)$ is undefined. That is, sample points such that $y-\mu(x)+\mu(x_i) \notin \Omega$. One should choose a small $k$; in numerical experiments we use $k=10$. We find the $k$ nearest sample points by querying a precomputed KD-tree \cite{kdtree} of all sample points. We check whether $y - \mu(x) + \mu(x_i) \in \Omega$ by querying a precomputed axis aligned bounding box tree (AABB tree) \cite{aabbtree} of the mesh cells used to discretize the problem. 
Per Section \ref{eq:local_mean_displacement_invariance}, $\Phi(y,x) \approx f_i$ when $x_i$ is close to $x$. The factor $\frac{V(x)}{V(x_i)}$, which was not discussed in Section \ref{eq:local_mean_displacement_invariance}, improves the approximation if $V(x)$ varies widely.

@@

at the point $x$ using radial basis function interpolation.

@@

In detail, we form $\widetilde{\Phi}(y,x) \approx \Phi(y,x)$ as follows:

@@

\widetilde{\Phi}(y,x) := \sum_{i=1}^k w_i~ b\left(\|x-x_i\|\right).

@@

Here $b_i$ are radial basis functions and $w_i$ are weights. The vector of weights, $w = (w_1, w_2, \dots, w_k)^T$, is found as the solution to the $k \times k$ linear system

@@

for $r>0$, and $b_i(0):=0$. For more details on radial basis function interpolation and polyharmonic splines, see \cite{rbf}.

@@

In the non-symmetric case, evaluating $\widetilde{\Phi}(y,x)$ requires us to perform one $k$-nearest neighbor query of the KD-tree of sample points, $k$ point collision queries of the AABB-tree of mesh cells, and solve a $k \times k$ linear system. In the symmetric case, we must perform two $k$-nearest neighbor queries of the KD-tree of sample points, $2k$ point collision queries of the AABB-tree of mesh cells, and solve a $4k \times 2k$ linear least squares problem via QR factorization. In both cases, this requires $O(k \log r + k \log N_\text{cell} + k^3)$ work, where $r$ is the number of sample points, and $N_\text{cell}$ is the number of cells in the mesh used to discretize the problem. Since $k$ is small, this is cheap.

@@

We convert $\mathbf{\Phi}$ into H1 matrix format using the standard geometrical clustering/adaptive cross method implemented within the HLIBpro software package \cite{HLIBPRO}. Although $\mathbf{\Phi}$ is a dense $\fedim \times \fedim$ matrix, constructing the H-matrix approximation only requires evaluation of $O(\hmatrixrank^2 \fedim \log \fedim)$ entries of $\mathbf{\Phi}$, and these entries are computed via interpolation procedure described in Section \ref{sec:approximate_kernel_entries}. Here $\hmatrixrank$ is the rank of the highest-rank block in the H-matrix. A dense representation of $\mathbf{\Phi}$ is never formed. We describe this H-matrix construction process in detail in Appendix \ref{app:h_matrix}. We convert $\bm{M}$ to H-matrix format using standard H-matrix methods for sparse matrices implemented within HLIBpro, using the same recursive block partitioning structure as was used for $\mathbf{\Phi}$.

@@

In detail, the process of constructing an H-matrix representation of $\AkerPcMat$ proceeds as follows. First, we construct hierarchical partitionings of the degrees of freedom for the columns and rows of the matrix (cluster trees, Section \ref{sec:cluster_trees}). Second, we construct a hierarchical partitioning of the blocks of the matrix, in such a way that many of the blocks in the partitioning are expected to be low rank, and the remaining high rank blocks are small (block cluster tree, Section \ref{sec:block_cluster_tree}). Finally, we form low rank approximations of the blocks of the matrix that are expected to be low rank (adaptive cross approximation, Section \ref{sec:adaptive_cross}), and fill in the remaining high rank  blocks with their numerical values. The first two steps require geometric information about the spatial locations of the degrees of freedom associated with the rows and columns of the matrix, but these steps do not depend on the particular values of matrix entries. The third step requires us to evaluate $O(\fedim \log \fedim)$ specially-chosen entries of $\AkerPcMat$, and we evaluate these entries using \eqref{eq:Akerpcmat_entries}.

@@

\begin{proof}[Proof of Theorem \ref{thm:vol_mean_cov}]
	Let $v \in L^2$ be arbitrary. We have
	\begin{align*}
		\left(v, \spatialvol\right)_{L^2(\Omega)} =& \int_\Omega v(x) \int_{\Omega} \left(\Aop \delta_x^*\right)^*(z) dz dx\\
		=& \int_\Omega v(x) \int_{\Omega} A(z,x) dz dx \\
		=& \int_\Omega \int_\Omega v(x) A(z,x) C(z) dz dx \\
		=& \left(\Aop^T C\right)(v),
	\end{align*}
	which implies $\spatialvol = \left(\Aop^T C\right)^*$.

>>>>

%The ellipsoid $E_x$ is not immediately available. 
To determine $E_x$, we must compute the scalar $\spatialvol(x)$, the length-$\gdim$ vector $\spatialmean(x)$, and the $\gdim \times \gdim$ matrix $\spatialcov(x)$.
%(recall $\gdim$ is the spatial dimension of $\Omega$).
The direct approach to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$ is to apply $\Aop$ to a point source centered at $x$ to get $\impulseresponse_x$, as per \eqref{eq:impulse_response_delta_action}. Then one can post-process $\impulseresponse_x$ to determine $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$. But this direct approach is computationally infeasible because it requires one operator action of $\Aop$ per point $x$ that we wish to determine $E_x$ at, and we will need to determine $E_x$ at a large number of points. Fortunately, it is possible to compute $\spatialvol(x)$, $\spatialmean(x)$, and $\spatialcov(x)$, \emph{for all points $x\in\Omega$ simultaneously}, by applying $\Aop^T$ to one constant function, $\gdim$ linear functions, and $\gdim(\gdim+1)/2$ quadratic functions. We present this efficient method of computing $\spatialvol$, $\spatialmean$, and $\spatialcov$ in Proposition \ref{thm:vol_mean_cov} and summarize the method for computing these moments in Algorithm \ref{alg:varhpi_mean_cov}.

\begin{prop}[Impulse response moments]
	\label{thm:vol_mean_cov}
%	Let $\spatialvol:\Omega \rightarrow \mathbb{R}$, $\spatialmean:\Omega \rightarrow \mathbb{R}^\gdim$, and $\spatialcov:\Omega \rightarrow \mathbb{R}^{\gdim \times \gdim}$, be the following functions:
%	\begin{equation*}
%	\spatialvol(x) := \int_{\Omega} \impulseresponse_x(y) dy, \qquad
%	\spatialmean(x) := \mathbb{E}(\widehat{\impulseresponse}_x), \qquad
%	\spatialcov(x) := \Var(\widehat{\impulseresponse}_x),
%	\end{equation*}
%	where $\widehat{\impulseresponse}_x := \impulseresponse_x / \spatialvol(x)$ denotes the normalized version of $\impulseresponse_x$, $\mathbb{E}(\widehat{\impulseresponse}_x)$ denotes the mean of $\widehat{\impulseresponse}_x$, and $\Var(\widehat{\impulseresponse}_x)$ denotes the variance of $\widehat{\impulseresponse}_x$.	
	Let  $x^i$ denote the $i^\text{th}$ component of $x$, i.e., $x = \left(x^1, x^2, \dots, x^\gdim\right)$. Let $C$, $\{L^i\}_{i=1}^\gdim$, and ${\{Q^{ij}\}_{i=1}^\gdim}_{j=1}^\gdim$ be the following constant, linear, and quadratic functions:
	\begin{equation*}
	C(x) := 1, \qquad
	L^i(x) := x^i, \qquad
	Q^{ij}(x) = x^i x^j.
	\end{equation*}
	Let $f/g$ denote the pointwise division of functions, $\left(f/g\right)(x) = f(x)/g(x)$, and $f\cdot g$ denote the pointwise multiplication of functions, $(f\cdot g)(x) = f(x)g(x)$. The following results hold:
	\begin{subequations}
		\label{eq:vol_mean_var}
		\begin{align}
		\spatialvol =& \left(\Aop^T C\right)^* \\
		\spatialmean^i =& \left(\Aop^T L^i\right)^* / \spatialvol \\
		\spatialcov^{ij} =& \left(\Aop^T Q^{ij}\right)^* / \spatialvol - \spatialmean^i\cdot \spatialmean^j
		\end{align}
	\end{subequations}
	for $i=1,\dots, \gdim$, $j=1,\dots,\gdim$.
\end{prop}

The proof of Proposition \ref{thm:vol_mean_cov} is shown in Appendix \ref{app:proofs}. The proof is straightforward, and follows directly from the definitions.

@@

%In Section \ref{sec:intro} we noted that computing $\Aker\left(y,x\right)$ is computationally costly. 
%This section describes how we compute approximate kernel entries
%\begin{equation*}
%\widetilde{\Phi}(y,x) \approx \Phi(y,x)
%\end{equation*}
%for arbitrary pairs of points $(y,x) \in \Omega \times \Omega$. 
Given $(y,x)\in \Omega \times \Omega$, let
\begin{equation}
\label{eq:fxyxp}
f_i := \frac{V(x)}{V(x_i)}\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right), \qquad i=1,\dots,k,
\end{equation}
%for arbitrary pairs of points $(y,x) \subset \Omega \times \Omega$.
% while Section \ref{sec:approximate_kernel_entries} we showed how we can cheaply evaluate approximate kernel entries at arbitrary pairs of points $(y,x) \in \Omega \times \Omega$.
where $\{x_i\}_{i=1}^k$ are the $k$ nearest sample points to $x$, excluding sample points for which $\phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right)$ is undefined. That is, sample points such that $y-\mu(x)+\mu(x_i) \notin \Omega$. One should choose a small $k$; in numerical experiments we use $k=10$. We find the $k$ nearest sample points by querying a precomputed KD-tree \cite{kdtree} of all sample points. We check whether $y - \mu(x) + \mu(x_i) \in \Omega$ by querying a precomputed axis aligned bounding box tree (AABB tree) \cite{aabbtree} of the mesh cells used to discretize the problem. 
Per the discussion in Section \ref{eq:local_mean_displacement_invariance}, if we ignore the factor $\frac{V(x)}{V(x_i)}$, then $\Phi(y,x) \approx f_i$ when $x_i$ is close to $x$. The factor $\frac{V(x)}{V(x_i)}$, which was not discussed in Section \ref{eq:local_mean_displacement_invariance}, is a form of preconditioning that improves this approximation if $V(x)$ varies widely.

@@

at the point $x$. This is done using the following radial basis function interpolation scheme:

@@

%In detail, we form $\widetilde{\Phi}(y,x) \approx \Phi(y,x)$ as follows:

@@

\Phi(y,x) \approx \widetilde{\Phi}(y,x) := \sum_{i=1}^k w_i~ b\left(\|x-x_i\|\right),

@@

where $b_i$ are radial basis functions and $w_i$ are weights. The vector of weights, $w = (w_1, w_2, \dots, w_k)^T$, is found as the solution to the $k \times k$ linear system

@@

for $r>0$, and $b_i(0):=0$. Polyharmonic spline radial basis function interpolation yields the smoothest interpolant of the given points and values, in a certain least-squares sense. For more details on radial basis function interpolation and polyharmonic splines, see \cite{rbf}.

@@

In the non-symmetric case, evaluating $\widetilde{\Phi}(y,x)$ requires us to perform one $k$-nearest neighbor query of the KD-tree of sample points, $k$ point collision queries of the AABB-tree of mesh cells, and solve a $k \times k$ linear system. In the symmetric case, we must perform two $k$-nearest neighbor queries of the KD-tree of sample points, $2k$ point collision queries of the AABB-tree of mesh cells, and solve a $4k \times 2k$ linear least squares problem via QR factorization. Both of these cases require $O(k \log r + k \log N_\text{cell} + k^3)$ work, where $r$ is the number of sample points, and $N_\text{cell}$ is the number of cells in the mesh used to discretize the problem. Since $k$ is small, this is cheap.

@@

We convert $\mathbf{\Phi}$ into H1 matrix format using the standard geometrical clustering/adaptive cross method implemented within the HLIBpro software package \cite{HLIBPRO}. Although $\mathbf{\Phi}$ is a dense $\fedim \times \fedim$ matrix, constructing the H-matrix approximation only requires evaluation of $O(\hmatrixrank^2 \fedim \log \fedim)$ entries $\mathbf{\Phi}_{ij} = \widetilde{\Phi}(p_i, p_j)$, and these entries are computed via the procedure described in Section \ref{sec:approximate_kernel_entries}. Here $\hmatrixrank$ is the rank of the highest-rank block in the H-matrix. A dense representation of $\mathbf{\Phi}$ is never formed. We describe this H-matrix construction process in detail in Appendix \ref{app:h_matrix}. We convert $\bm{M}$ to H-matrix format using standard H-matrix methods for sparse matrices implemented within HLIBpro, using the same recursive block partitioning structure as was used for $\mathbf{\Phi}$.

@@

The process of constructing an H-matrix representation of $\mathbf{\Phi}$ proceeds as follows. First, we construct hierarchical partitionings of the degrees of freedom for the columns and rows of the matrix (cluster trees, Section \ref{sec:cluster_trees}). Second, we construct a hierarchical partitioning of the blocks of the matrix, in such a way that many of the blocks in the partitioning are expected to be low rank, and the remaining high rank blocks are small (block cluster tree, Section \ref{sec:block_cluster_tree}). Finally, we form low rank approximations of the blocks of the matrix that are expected to be low rank (adaptive cross approximation, Section \ref{sec:adaptive_cross}), and fill in the remaining high rank  blocks with their numerical values. The first two steps require geometric information about the spatial locations of the degrees of freedom associated with the rows and columns of the matrix, but these steps do not depend on the particular values of matrix entries. The third step requires us to evaluate $O(\fedim \log \fedim)$ specially-chosen entries of $\AkerPcMat$, and we evaluate these entries using \eqref{eq:Akerpcmat_entries}.

@@

\begin{proof}[Proof of Proposition \ref{thm:vol_mean_cov}]
	Let $v \in L^2$ be arbitrary. We have
	\begin{align*}
		\left(v, \spatialvol\right)_{L^2(\Omega)} =& \int_\Omega v(x) \int_{\Omega} \left(\Aop \delta_x^*\right)^*(z) dz dx\\
		=& \int_\Omega v(x) \int_{\Omega} A(z,x) dz dx \\
		=& \int_\Omega \int_\Omega v(x) A(z,x) C(z) dz dx \\
		=& \left(\Aop^T C\right)(v),
	\end{align*}
	which implies $\spatialvol = \left(\Aop^T C\right)^*$.
